Temporal graph realization from fastest paths

### 2023-12-20 STACS2023/rebuttal_response.txt

We thank all of the reviewers for their detailed analysis of our work. We will take in consideration all of the remarks and incorporate them to our paper before the final submission. We now provide a short response to each of the reviewers separately.

REVIEWER 1
We appreciate the reviewer's thoughts and would like to gently clarify an aspect of our research problem. 
We are focused on the fastest temporal paths, rather than the shortest paths. Fastest temporal paths can be thought of as an extension of the concept of shortest paths on (static) graphs, to the setting of temporal graphs, where due to causality, entities and information can "flow" only along sequences of edges whose time-labels are strictly increasing.
In our context, a fastest temporal path from vertex u to v is defined as the temporal path that uses the smallest possible duration to travel from u to v. This "duration" is essentially the "travel time" from the starting vertex to the ending one. In our example in Figure 1, the journey from v_1 to v_k can begin at time 7 and finish a time 21, which results in a travel time of 15 units and not 21. The initial six units of time before the journey do not contribute to the travel time. To put it into a real-life scenario, consider the example of travelling: we normally do not start measuring the travel duration at 00:00 on the day of travel, but at the specific time we live our home.

REVIEWER 2

#MINOR REMARKS
- l 230: We will remove the requirement, that every vertex has at least one neighbor of every color. This was accidentally kept from an earlier draft of the reduction, and is not needed in the reduction presented in the paper.
- Figure 5: One of the degree one neighbors of the bottom left green vertex should be in Z^*, we made a mistake.
- l 505: The reviewer is right, segment is unique whenever it does not consist of a single edge. We will add this detail. Note that our algorithm remains correct, since there can be at most 1 (constant) such segment between a pair of important vertices.

REVIEWER 3

#MAJOR REMARKS
1)
The reviewer is correct. In our W[1]-hardness proof we do not explicitly describe the values of the D matrix for all pairs of vertices. We apologise for this oversight and acknowledge that we failed to mention a crucial remark, that the values for durations of fastest paths among non-adjacent vertex pairs, that we do not explicitly specify in the proof, are set to infinity. We will clarify this in the updated version of the paper (in the full version this is mentioned - see Appendix in lines 563-564).

2)
When discussing the FPT algorithm, we use the term "guessing" where we mean that we enumerate all possible outcomes. We will state this more clearly in the next iteration of the paper.

3)
At the start of the algorithm, we chose a spanning tree T of G' and use it to determine the set of feedback edges F and set(s) of important vertices. We then "guess" the fastest paths among the determined important vertices in the whole graph G'. This implies that the fastest paths use the edges of our spanning tree T and the feedback edges F. Therefore, the choice of a spanning tree (and consequently the feedback edges F) does not influence the structure of fastest paths.

#MINOR REMARKS
Considering the graph realization on temporal graphs we agree that there are many different settings one could study, and for each of them argue why it is interesting.
A nice property in periodic temporal graphs is that a temporal path among each pair of vertices exists (whenever the underlying graph is connected), which is not necessarily true in the non-periodic case. This is why we chose to study the problem on periodic temporal graphs.
Initiating the study we tried to create the simplest possible setting, which resulted in labeling each edge with exactly one label in a fixed time period.
We do state that the periodic case with all edges having the same period is the most fundamental case, where the more evolved variant would be graphs with the edges admitting different periods.
As the reviewer mentioned, there are also other possible definitions of temporal paths that would be worth exploring, such as the shortest or foremost paths, both with the increasing or non-decreasing time labels. We focused on the fastest temporal paths with strictly increasing time labels, as we believe they are the most natural in our setting.

Answering the questions about the FPT algorithm. 
- To guess a fastest path among two vertices we produce f(k) guesses. Thus guessing a path among n^2 pairs of vertices would result in f(k)^{n^2} guesses, which is not an FPT running time anymore. 
We believe this explanation answers also the reasoning behind the number of guesses in lines 496-498. 
- Let us shortly outline why the set Z^* is necessary. Let u be a clip vertex of a tree T_u, w \in V(T_u) a vertex from Z^*, and v an arbitrary vertex from the graph. Suppose that P is a fastest temporal path from v to u, of duration d that reaches u at time 1 (mod delta), and let Q be another (v,u)-temporal path that reaches u at time 5 (modulo delata) and is of duration d+1.
Suppose also that the edge uw is labeled with 6. Then the (u,w)-temporal path R_1 = Q \cup {uw} is of smaller duration than R_2 = P \cup {uw}. Which implies that knowing the fastest temporal path from v to u does not provide enough information to uniquely determine the fastest temporal path from v to w (or to any other vertex from T_u).
Now observe that knowing the label of one edge of the tree T_u results in uniquely determining all other labels, because of the fact that there is a unique (static) path between any two vertices of T_u (we can use our tree algorithm for this part).
-line 506: one path is indeed reverse of the other.

### 2023-12-20 STACS2023/STACS23-shortversion.tex
dec 20

\problemdef{\deltaExactLong\ (\deltaExact)}
{An $n \times n$ matrix $D$, a positive integer $\Delta$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is $D_{i,j}$?}

\begin{figure}[t]
    \centering
	\begin{tikzpicture}[xscale=1.5]
		%%%S_uv
		\node[vert,label=below:$v_1$] (1) at (1,0) {};
		\node[vert,label=below:$v_2$] (2) at (3,0) {};
		\node[vert,label=below:$v_3$] (3) at (5,0) {};
		\node[vert,label=below:$v_4$] (4) at (7,0) {};
		\node[vert,label=below:$v_5$] (5) at (9,0) {};
		\draw (1) -- node[label=above:$10t+7$] {} (2) -- node[label=above:$10t+3$] {}  (3) -- node[label=above:$10t+5$] {} (4) -- node[label=above:$10t+1$] {} (5);	
	\end{tikzpicture}
	\caption{An example of a $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$, where $\Delta = 10$ and the 10-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,10\}$ is as follows: $\lambda(v_1 v_2)=7$, $\lambda(v_2 v_3)=3$, $\lambda(v_3 v_4)=5$, and $\lambda(v_4 v_5)=1$. 
    Here, the fastest temporal path from $u$ to $v$ traverses the first edge $v_1v_2$ at time $7$, second edge $v_2v_3$ a time $13$, third edge $v_3v_4$ at time $15$ and the last edge $v_4v_5$ at time $21$.
    This results in the total duration of $15$ for the fastest temporal path from $v_1$ to $v_5$.
	\label{fig:periodic-example}}
\end{figure}

@@

 We have a \emph{verification gadget} for every color $i$. They interact with the edge selection gadgets as follows.
 The verification gadget for color $i$ is connected to all edge selection gadgets that involve color $i$. More specifically, this is connected to every path corresponding to an edge at a position in the path that encodes the endpoint of color $i$ of that edge (again, see \cref{fig:hardness1}) for reference. Intuitively, the distances in the verification gadget are only realizable if the selected edges all have the same endpoint of color $i$.
Hence, the distances of all verification gadgets can be realized if and only if the selected edges form a clique.

@@

Before presenting the structure of our algorithm for \cref{thm:FPTwrtFES}, observe that, in a static graph, the number of paths between two vertices can be upper-bounded by a function $f(k)$ of the feedback edge number $k$ of the graph. 
Therefore, for any fixed pair of vertices $u$ and $v$, we can ``guess'' the edges of the fastest temporal path from $u$ to $v$. 
However, 
% due to the difficulties illustrated in~\cref{fig:ftpExample}, 
for an FPT algorithm with respect to $k$, we cannot afford to guess the edges of the fastest temporal path for each of the $O(n^2)$ pairs of vertices. 
To overcome this difficulty, our algorithm follows this high-level strategy:
\begin{itemize}
    %%%\item We perform a preprocessing step by iteratively removing all vertices of degree one.
    \item We identify a small number $f(k)$ of ``important vertices''; these consist of the sets that we call $U, U^{\ast}, Z^{\ast}$.
    \item For each pair $u,v$ of important vertices, we guess the edges of the fastest temporal path from $u$ to $v$ (and from $v$ to $u$).
    \item From these guesses we can still not deduce the edges of the fastest temporal paths between many pairs of non-important vertices. However, as we prove, it suffices to guess only a small number of specific auxiliary structures (to be defined later).
    \item From these guesses we deduce fixed relationships between the labels of most of the edges of the graph. 
    \item For all the edges, for which we do not have deduced a label yet, we introduce a \emph{variable}. Using all these variables, we build an Integer Linear Program (ILP). 
    Among the constraints in this ILP we have that, for each of the $O(n^2)$ pairs of vertices $u,v$ in the graph, the duration of one specific temporal path from $u$ to $v$ (according to our guesses) is \emph{equal} to the desired duration $D_{u,v}$, while the duration of each of the other temporal paths from $u$ to $v$ is \emph{at least} $D_{u,v}$.
    \item By making any of the above guesses, we restrict the solution space for the problem \deltaExact. This restricted solution space coincides with the set of feasible solutions to the resulting ILP. Furthermore, the set of feasible solutions for all constructed ILPs coincide with the set of all solutions to \deltaExact\ (\ie regardless of our guesses). As each ILP can be solved in FPT time with respect to $k$ by Lenstra's Theorem~\cite{Lenstra1983Integer} (the number of variables is upper bounded by a function of $k$), we obtain our FPT algorithm for \deltaExact\ with respect to $k$.
\end{itemize}

@@

\subparagraph{Important vertices.}
Let $D$ be the input matrix of \deltaExact\, and let $G$ be its underlying graph, on $n$ vertices and $m$ edges.
%\subsubsection*{Preprocessing of the input}
From the underlying graph $G$ of $D$ we first create a graph $G'$ by
iteratively removing vertices of degree one from $G$,
and denote with 
$
Z = V(G) \setminus V(G')
$, the set of removed vertices.
%Then we determine a minimum feedback edge set $F$ of $G'$.
%by finding a spanning tree $T$ of $G'$ and set $F = E(G) \setminus E(T)$. 
%Note that $F$ is also a minimum feedback edge set of $G$.
Then we determine the set $U$ (the \emph{``vertices of interest''}), and the set $U^*$ (the neighbors of the vertices of interest), as follows.
Let $T$ be a spanning tree of $G'$, with $F$ being the corresponding feedback edge set of $G'$.
Let $V_1 \subseteq V(G')$ be the set of leaves in the spanning tree $T$, $V_2 \subseteq V(G')$ be the set of vertices of degree two in $T$ which are incident to at least one edge in $F$, 
and let $V_3 \subseteq V(G')$ be the set of vertices of degree at least $3$ in $T$. 
Then $|V_1| + |V_2| \leq 2k$, since every leaf in $T$ and every vertex in $V_2$ is incident to at least one edge in $F$,
and $|V_3| \leq |V_1|$ by the properties of trees.
We denote with 
\[U = V_1 \cup V_2 \cup V_3\]
the set of \emph{vertices of interest}. It follows that $|U| \leq 4k$.
We set $U^*$ to be the set of vertices in $V(G') \setminus U$ that are neighbors of vertices in $U$, \ie 
\[U^* = \{v \in V(G') \setminus U  :  u \in U, v \in N(u)\}.\]
Again, using the tree structure, we get that for any $u \in U$ its neighborhood is of size $|N(u)| \in O(k)$, since every neighbor of $u$ is the first vertex of a (unique) path to another vertex in $U$.
It follows that $|U^*| \in O(k^2)$.
From the construction of $Z$ (by iteratively removing vertices of degree one from $G$)
it follows that $Z$ consists of disjoint trees $T_1, T_2, \dots$.
For a tree $T_i$ we denote with $u_i$ the vertex in $G'$ that is a neighbor of a vertex in $T_i$, 
and call it a \emph{clip vertex of the tree $T_i$}.
It follows that there can be many different trees $T_i$ that are incident to the same clip vertex $u_i \in V(G')$,
but each tree $T_i$ is incident to exactly one clip vertex $u_i \in V(G')$.
Since $u_i$ is the only vertex connecting all of the trees $T_i$ incident to it,
from now on we assume that a tree $T_{u_i}$ in $Z$ is a union of trees on vertices from $V(G) \setminus V(G')$,
that are clipped at the same vertex $u_i \in V(G')$.
For each of the trees $T_{u_i}$ in $Z$, we select one vertex $r_i$, that is a neighbor of the clip vertex $u_i$,
and call it \emph{a representative vertex of the tree $T_{u_i}$}.
We now define as $Z^*$ the set of representatives $r_i$ of trees $T_i \in Z$,
where the clip vertex $v_i$ of $T_i$ is a vertex of interest, \ie
\[
Z^* = \{ r_i  :  r_i \in T_i, \text{ where $T_i \in Z$, the clip vertex $u_i$ of $T_i$ is in $U$, and $r_i u_i \in E(G)$}\}.
\]
Since there are $O(k)$ vertices of interest, we get that $|Z^*| \in O(k)$.
Finally, the set of \emph{important vertices} is defined as the set $U \cup U^{\ast} \cup Z^{\ast}$.
For an illustration see \cref{fig:labelingVertices}.

>>>>

\problemdef{\deltaExactLong\ (\deltaExact)}
{An integer $n \times n$ matrix $D$, a positive integer $\Delta$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is $D_{i,j}$?}

\begin{figure}[t]
    \centering
	\begin{tikzpicture}[xscale=1.5]
		%%%S_uv
		\node[vert,label=below:$v_1$] (1) at (1,0) {};
		\node[vert,label=below:$v_2$] (2) at (3,0) {};
		\node[vert,label=below:$v_3$] (3) at (5,0) {};
		\node[vert,label=below:$v_4$] (4) at (7,0) {};
		\node[vert,label=below:$v_5$] (5) at (9,0) {};
		\draw (1) -- node[label=above:$10t+7$] {} (2) -- node[label=above:$10t+3$] {}  (3) -- node[label=above:$10t+5$] {} (4) -- node[label=above:$10t+1$] {} (5);	
	\end{tikzpicture}
	\caption{An example of a $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$, where $\Delta = 10$ and the 10-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,10\}$ is as follows: $\lambda(v_1 v_2)=7$, $\lambda(v_2 v_3)=3$, $\lambda(v_3 v_4)=5$, and $\lambda(v_4 v_5)=1$. 
    Here, the fastest temporal path from $v_1$ to $v_2$ traverses the first edge $v_1v_2$ at time $7$, second edge $v_2v_3$ a time $13$, third edge $v_3v_4$ at time $15$ and the last edge $v_4v_5$ at time $21$.
    This results in the total duration of $21 - 7 + 1 = 15$ for the fastest temporal path from $v_1$ to $v_5$.
	\label{fig:periodic-example}}
\end{figure}

@@

 We have a \emph{verification gadget} for every color $i$. They interact with the edge selection gadgets as follows.
 The verification gadget for color $i$ is connected to all edge selection gadgets that involve color $i$. More specifically, this is connected to every path corresponding to an edge at a position in the path that encodes the endpoint of color $i$ of that edge (again, see \cref{fig:hardness1} for reference). Intuitively, the distances in the verification gadget are only realizable if the selected edges all have the same endpoint of color $i$.
Hence, the distances of all verification gadgets can be realized if and only if the selected edges form a clique.

@@

Before presenting the structure of our algorithm for \cref{thm:FPTwrtFES}, observe that, in a static graph, the number of paths between two vertices can be upper-bounded by a function $f(k)$ of the feedback edge number $k$ of the graph. 
Therefore, for any fixed pair of vertices $u$ and $v$, we can ``guess'' the edges of the fastest temporal path from $u$ to $v$. 
However, 
% due to the difficulties illustrated in~\cref{fig:ftpExample}, 
for an FPT algorithm with respect to $k$, we cannot afford to guess the edges of the fastest temporal path for each of the $O(n^2)$ pairs of vertices. 
To overcome this difficulty, our algorithm follows this high-level strategy:
\begin{itemize}
    %%%\item We perform a preprocessing step by iteratively removing all vertices of degree one.
    \item We identify a small number $f(k)$ of ``important vertices''; these consist of the sets that we call $U, U^{\ast}, Z^{\ast}$.
    \item For each pair $u,v$ of important vertices, we guess the edges of the fastest temporal path from $u$ to $v$ (and from $v$ to $u$).
    \item From these guesses we can still not deduce the edges of the fastest temporal paths between many pairs of non-important vertices. However, as we prove, it suffices to guess only a small number of specific auxiliary structures (to be defined later).
    \item From these guesses we deduce fixed relationships between the labels of most of the edges of the graph. 
    \item For all the edges, for which we do not have deduced a label yet, we introduce a \emph{variable}. Using all these variables, we build an Integer Linear Program (ILP). 
    Among the constraints in this ILP we have that, for each of the $O(n^2)$ pairs of vertices $u,v$ in the graph, the duration of one specific temporal path from $u$ to $v$ (according to our guesses) is \emph{equal} to the desired duration $D_{u,v}$, while the duration of each of the other temporal paths from $u$ to $v$ is \emph{at least} $D_{u,v}$.
    \item 
    %%%By making any of the above guesses, we restrict the solution space for the problem \deltaExact. This restricted solution space coincides with the set of feasible solutions to the resulting ILP. Furthermore, the set of feasible solutions for all constructed ILPs coincide with the set of all solutions to \deltaExact\ (\ie regardless of our guesses). 
    By making each of the above combinations of guesses, we essentially enumerate all possible ways that our instance of \deltaExact\ has a solution, and for each of these possible ways we create an ILP. That is, our instance of \deltaExact\ has a solution if and only if at least one of these ILPs has a feasible solution.
    As each ILP can be solved in FPT time with respect to $k$ by Lenstra's Theorem~\cite{Lenstra1983Integer} (the number of variables is upper bounded by a function of $k$), we obtain our FPT algorithm for \deltaExact\ with respect to $k$.
\end{itemize}

@@

\subparagraph{Important vertices.}
Let $D$ be the input matrix of \deltaExact, and let $G$ be its underlying graph, on $n$ vertices and $m$ edges.
%\subsubsection*{Preprocessing of the input}
From the underlying graph $G$ of $D$ we first create a graph $G'$ by
iteratively removing vertices of degree one from $G$,
and denote with 
$
Z = V(G) \setminus V(G')
$, the set of removed vertices.
%Then we determine a minimum feedback edge set $F$ of $G'$.
%by finding a spanning tree $T$ of $G'$ and set $F = E(G) \setminus E(T)$. 
%Note that $F$ is also a minimum feedback edge set of $G$.
Then we determine the set $U$ (the \emph{``vertices of interest''}), and the set $U^*$ (the neighbors of the vertices of interest), as follows.
Let $T$ be a spanning tree of $G'$, with $F$ being the corresponding feedback edge set of $G'$.
Let $V_1 \subseteq V(G')$ be the set of leaves in the spanning tree $T$, $V_2 \subseteq V(G')$ be the set of vertices of degree two in $T$ which are incident to at least one edge in $F$, 
and let $V_3 \subseteq V(G')$ be the set of vertices of degree at least $3$ in $T$. 
Then $|V_1| + |V_2| \leq 2k$, since every leaf in $T$ and every vertex in $V_2$ is incident to at least one edge in $F$,
and $|V_3| \leq |V_1|$ by the properties of trees.
We denote with 
\[U = V_1 \cup V_2 \cup V_3\]
the set of \emph{vertices of interest}. It follows that $|U| \leq 4k$.
We set $U^*$ to be the set of vertices in $V(G') \setminus U$ that are neighbors of vertices in $U$, \ie 
\[U^* = \{v \in V(G') \setminus U  :  u \in U, v \in N(u)\}.\]
Again, using the tree structure, we get that for any $u \in U$ its neighborhood is of size $|N(u)| \in O(k)$, since every neighbor of $u$ is the first vertex of a (unique) path to another vertex in $U$.
It follows that $|U^*| \in O(k^2)$.
From the construction of $Z$ (by iteratively removing vertices of degree one from $G$)
it follows that $Z$ consists of disjoint trees $T_1, T_2, \dots$.
For a tree $T_i$ we denote with $u_i$ the vertex in $G'$ that is a neighbor of a vertex in $T_i$, 
and call it a \emph{clip vertex of the tree $T_i$}.
It follows that there can be many different trees $T_i$ that are incident to the same clip vertex $u_i \in V(G')$,
but each tree $T_i$ is incident to exactly one clip vertex $u_i \in V(G')$.
Since $u_i$ is the only vertex connecting all of the trees $T_i$ incident to it,
from now on we assume that a tree $T_{u_i}$ in $Z$ is a union of trees on vertices from $V(G) \setminus V(G')$,
that are clipped at the same vertex $u_i \in V(G')$.
For each of the trees $T_{u_i}$ in $Z$, we select one vertex $r_i$, that is a neighbor of the clip vertex $u_i$,
and call it \emph{a representative vertex of the tree $T_{u_i}$}.
We now define as $Z^*$ the set of representatives $r_i$ of trees $T_i \in Z$,
where the clip vertex $v_i$ of $T_i$ is a vertex of interest, \ie
\[
Z^* = \{ r_i  :  r_i \in T_i, \text{ where $T_i \in Z$, the clip vertex $u_i$ of $T_i$ is in $U$, and $r_i u_i \in E(G)$}\}.
\]
Since there are $O(k)$ vertices of interest, we get that $|Z^*| \in O(k)$.
Finally, the set of \emph{important vertices} is defined as the set $U \cup U^{\ast} \cup Z^{\ast}$.
For an illustration see \cref{fig:labelingVertices}.
